---
title: "STOCfree: prediction of probabilities of freedom from infection from longitudinal data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim of the STOCfree package is predict herd level probabilities of freedom from infection from longitudinal data collected as part of surveillance programmes.

Below, the use of the package is presented through the analysis of a small dataset collected for the surveillance of infection by the BVDV virus in cattle.

# Package installation

The easiest way to install the `STOCfree` package is from Github. This requires installing the `devtool` package first.

```{r, eval = FALSE, message=FALSE}
install.packages("devtools")
```

Then load the `devtool` package:

```{r, eval = FALSE}
library(devtools)
```

and install the STOCfree package

```{r, eval = FALSE}
install_github("AurMad/STOCfree")
```

# Loading the STOCfree package

```{r}
library(STOCfree)
```

# Data

The `STOCfree` package contains a dataset called `herdBTM` which contains the results of antibody ELISA tests performed on bulk tank milk. Each row is a testing date in a herd. There are 100 herds with 11 tests for each herd.

```{r}
head(herdBTM)
```

# Formatting the data for analysis

Herds in the `herdBTM` are tested approximately every 6 months. The STOCfree model models infection with a montlhy time step. The data used by the model need to have one row per month. The herdBTM data is expanded to have one row per month with the `expand_month()` function.

```{r, message=FALSE}
herdBTM_month <- expand_month(data = herdBTM,
                              herd_colname = Farm,
                              date_colname = DateOfTest,
                              test_res_colname = TestResult)

herdBTM_month
```

# Priors

## Test

Prior for tests are stored in a variable called `test_priors`.

```{r}
test_priors <- list(
  Se_beta_a = 12,
  Se_beta_b = 2,
  Sp_beta_a = 200,
  Sp_beta_b = 4
)

```

## Infection dynamics

Probability of being infected on the first testing time for a herd and probability of not eiliminating the infection between 2 consecutive tests.

```{r}
infection_priors <- list(
  pi1_beta_a = 1,
  pi1_beta_b = 2,
  tau2_beta_a = 30,
  tau2_beta_b = 2
 )
```

and priors for risk factors

```{r}
risk_factor_priors <- list(
  theta_norm_mean = 0,
  theta_norm_sd = .01
)
```



# STOC free model

```{r}
library(rjags)
```


## Compiling

The JAGS model is compiled using the `compile_JAGS()` function.

```{r}
test <- expand_month(data = herdBTM[herdBTM$Farm %in% c("FR001", "FR002"),],
                     herd_colname = Farm,
                     date_colname = DateOfTest,
                     test_res_colname = TestResult)

compiled_model <- compile_JAGS(test_data = test, 
             herd_id = herd_id, 
             row_id = row_id,
             month = month_id,
             risk_factors = c("ln_nOrig6_12", "LocalSeroPrev"),
             test_result = TestResult,
             test_priors = test_priors, 
             infection_priors = infection_priors, 
             risk_factor_priors = risk_factor_priors,
             n_chains = 2)
```

## Burnin


```{r}
update(compiled_model, n.iter = 100)
```


## Sampling

```{r}
savedParam <- c("Se", "Sp", "theta", "tau2", "piPred", "StatusPred")

model_samples <- rjags::coda.samples(compiled_model,
                            variable.names = savedParam,
                            n.iter = 1000,
                            thin = 5)
```


## Results
