---
title: "STOCfree: prediction of probabilities of freedom from infection from longitudinal data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim of the STOCfree package is to predict herd level probabilities of freedom from infection from longitudinal data collected as part of surveillance programmes.

Below, the use of the package is presented through the analysis of a small dataset collected for the surveillance of infection by the BVDV virus in cattle.

# Package installation

Before installing the package, you need to make sure that JAGS is installed. This programme can be installed from the following website: https://sourceforge.net/projects/mcmc-jags/files/.

The easiest way to install the `STOCfree` package is from Github. This requires installing the `devtool` package first.

```{r, eval = FALSE, message=FALSE}
install.packages("devtools")
```

Then load the `devtool` package:

```{r, eval = FALSE}
library(devtools)
```

and install the STOCfree package:

```{r, eval = FALSE}
install_github("AurMad/STOCfree")
```

# Attaching the STOC free package

The `STOCfree` package needs to be attached.

```{r}
library(STOCfree)
```

# Data

The `STOCfree` package contains a dataset called `herdBTM` which contains the results of antibody ELISA tests performed on bulk tank milk. Each row is a testing date in a herd. There are 100 herds with 11 tests for each herd.

```{r}
head(herdBTM)
```

# Formatting the data for analysis

Herds in the `herdBTM` are tested approximately every 6 months. The STOCfree model models infection with a montlhy time step. The data used by the model need to have one row per month. The herdBTM data is expanded to have one row per month with the `expand_month()` function.

```{r, message=FALSE}
herdBTM_month <- expand_month(data = herdBTM,
                              herd_colname = Farm,
                              date_colname = DateOfTest,
                              test_res_colname = TestResult)

herdBTM_month
```

# Priors

## Test

Prior for tests are stored in a variable called `test_priors`.

```{r}
test_priors <- list(
  Se_beta_a = 12,
  Se_beta_b = 2,
  Sp_beta_a = 200,
  Sp_beta_b = 4
)

```

## Infection dynamics

Probability of being infected on the first testing time for a herd and probability of not eliminating the infection between 2 consecutive tests.

```{r}
infection_priors <- list(
  pi1_beta_a = 1,
  pi1_beta_b = 2,
  tau2_beta_a = 30,
  tau2_beta_b = 2
 )
```

and priors for risk factors

```{r}
risk_factor_priors <- list(
  theta_norm_mean = 0,
  theta_norm_sd = .01
)
```

# STOC free model

## Compiling

The JAGS model is compiled using the `compile_JAGS` function. We just use 3 herds to test the code.

```{r}
test <- expand_month(data = herdBTM[herdBTM$Farm %in% c("FR001", "FR002", "FR003"),],
                     herd_colname = Farm,
                     date_colname = DateOfTest,
                     test_res_colname = TestResult)

compiled_model <- compile_JAGS(test_data = test, 
             herd_id = herd_id, 
             row_id = row_id,
             month = month_id,
             risk_factors = c("ln_nOrig6_12", "LocalSeroPrev"),
             test_result = TestResult,
             test_priors = test_priors, 
             infection_priors = infection_priors, 
             risk_factor_priors = risk_factor_priors,
             n_chains = 2)
```

## Samples from the parameter posterior distributions

Samples from the posterior distributions of model parameters, predicted probabilities of infection and predicted statuses are drawn using the `sample_model` function. 

```{r}
samples <- sample_model(compiled_model, n_burnin = 100, n_iter = 100, n_thin = 5)
```

# Results

The model retruns a list with 3 components:

- samples from the model parameters posterior distributions
- samples from the predicted probability of infection posterior distributions
- predicted model performance

## Model parameters

The samples from the model parameter posterior distributions are stored in the `parameters` component of the variable in which we have stored the model results

```{r}
param <- samples$parameters
```

## Probability of infection

The samples from the model predicted probability of infection posterior distributions are stored in the `proba_inf` component of the variable in which we have stored the model results

```{r}
proba_inf <- samples$proba_inf
```


## Model performance

Needs a better description

```{r}
cutoff <- samples$choice_cutoff
```

